---
title: "p22162_Lara_Analysis_v2"
author: "DTG"
date: '`r Sys.Date()`'
output: 
  rmdformats::robobook:
    fig_width: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE, 
                      fig.width = 8, fig.height = 4,
                      cache=FALSE, cache.lazy = FALSE, cache.rebuild = FALSE)
library(tidyverse)
library(ggforce)
library(grid)
library(ggpubr)
library(yaml)
library(DESeq2)
library(EDASeq)
library(DT)
#library(shiny)
library(EnhancedVolcano)
library(tools)
library(ComplexHeatmap)
library(gridExtra)
library(gtable)
library(circlize)
library(openxlsx)
library(reshape2)
library(kableExtra)
library(rlang)
library(forcats)
library(gtools)
library(RColorBrewer)
library(fgsea)
library(MatrixGenerics)
#library(plyr)

counts <- DESeq2::counts
#config <- yaml.load_file("config.yaml")
source("functions.R")
```

<style type="text/css">
.book .book-body .page-inner {
  max-width: 1600px;
  margin-left: auto;
  margin-right: auto;
}
</style>

## Study overview

```{r reset_analysis_obj, eval=TRUE, include=FALSE}
# This is moved out of the setup block to avoid accidentally overwriting your
# analysis object when restarting a session
analysis = list()
```

```{r load_config, include=FALSE}
# Store config with results object for record keeping
analysis$config <- yaml.load_file(here("config/config_v2.yml"))
```

```{r sample_metadata_setup}

analysis$sampleTable <- read.table(here("config/p22162_Lara_exp_design.txt"),
                                   header=T, stringsAsFactors = FALSE, 
                                   sep="\t", quote='"')
analysis$sampleTable$CollectionDate <- as.Date(analysis$sampleTable$CollectionDate,
                                               tryFormats=c("%m/%d/%y"))
analysis$sampleTable$CollectionSite <- plyr::mapvalues(analysis$sampleTable$CollectionSite,
                                                       from=c(1,2), to=c('Bladder', 'Other'))
analysis$sampleTable$TumorRegions <- as.integer(analysis$sampleTable$TumorRegions)
analysis$sampleTable$Age <- as.integer(analysis$sampleTable$Age)
analysis$sampleTable$Batch <- as.factor(analysis$sampleTable$Batch)
knitr::kable((analysis$sampleTable %>%
                select(c(SampleID, CollectionDate, Age, TumorRegions,
                         MatchedNormal,CollectionSite, Group, Batch))),
             position='center', table.attr="style='width:100%;'", format='html') %>% 
  kable_styling(full_width = TRUE)
analysis$sampleTable <- analysis$sampleTable %>% 
  dplyr::arrange(across(all_of(analysis$config$sampleGrouping)))
#analysis$sampleTable <- as.data.frame(sapply(analysis$sampleTable, as.factor))


# Store the order you'd like samples to appear in figure
# Order is determined by sampleGrouping parameters in config
# But you could also manually specify it here if there aren't easy rules
sample_order <- fct_inorder(as.character(analysis$sampleTable$SampleID))
```

```{r}
if (!is.null(unlist(analysis$config$dropSamples))){
  out <- paste0("Sample(s)\n", 
                paste0(analysis$config$dropSamples, sep=", ", collapse = ''),
                "\nbeing removed from downstream analysis per user specification.") 
  analysis$sampleTable <- analysis$sampleTable %>% 
    filter(!SampleID %in% analysis$config$dropSamples)
  sample_order <- sample_order[!(sample_order %in% as.factor(analysis$config$dropSamples))]
  cat(out)
} else {
  analysis$ddsDrop <- analysis$dds
}
sample_order <- droplevels(sample_order)
```

```{r read_counts_tables, include=FALSE}
analysis$samplefileIDs <- analysis$sampleTable$FileID

analysis$sampleSTARReads <- sapply(analysis$sampleTable$FileID, function(sid) {
  read_tsv(file.path(analysis$config$rootDir,
                     analysis$config$alignmentDir,paste0(sid,analysis$config$STARreadSuffix)),
  col_names = c("gene_id","unstranded_count","sense_count","antisense_count"),
  col_types = c(gene_id = col_character(),
  unstranded_count = col_double(),
  sense_count = col_double(),
  antisense_count = col_double()))},
simplify = FALSE,
USE.NAMES = TRUE)
```

```{r dds, include=FALSE}
analysis$mapBins <- do.call(cbind,
                            sapply(analysis$sampleTable$FileID, function(sid){
                              analysis$sampleSTARReads[[sid]][,paste0(analysis$config$libraryType,"_count")][c(1:4),]}))
rownames(analysis$mapBins) <- analysis$sampleSTARReads[[1]][c(1:4),][["gene_id"]]
colnames(analysis$mapBins) <- recode(as.character(unlist(strsplit(analysis$sampleTable$SampleID, paste0(".", analysis$config$libraryType, "_count")))), !!!setNames(analysis$sampleTable$SampleID, as.character(analysis$sampleTable$FileID)))
 
raw_counts <- do.call(cbind,sapply(analysis$sampleTable$FileID,
                                   function(sid){
                                     analysis$sampleSTARReads[[sid]][,paste0( analysis$config$libraryType, "_count")][-c(1:4),]}))
colnames(raw_counts) <- recode(as.character(unlist(strsplit(analysis$sampleTable$FileID,
                                                            paste0(".", analysis$config$libraryType, "_count")))),
                               !!!setNames(analysis$sampleTable$SampleID,
                                           as.character(analysis$sampleTable$FileID)))
rownames(analysis$sampleTable) <- analysis$sampleTable$SampleID
analysis$sampleTable <- analysis$sampleTable[colnames(raw_counts),]
analysis$dds <- DESeqDataSetFromMatrix(raw_counts,
  analysis$sampleTable,
  design = as.formula(paste0("~ ", analysis$config$design))
)
analysis$mapBins <- rbind(analysis$mapBins,"N_identified" = colSums2(DESeq2::counts(analysis$dds)))
```

## QC {.tabset}

### Mapping

Mapping outcomes of reads from each sample. Reads need to be 'identified' or successfully mapped to a single feature to be used for gene expression analysis.

```{r mapping_plot, fig.width=12}
mapping_plot(analysis$mapBins, sort=TRUE)
```

```{r gene_names, include=FALSE}
geneInfoTab <- read_tsv(file.path(analysis$config$rootDir,
                                  analysis$config$referenceDir, "geneInfo.tab"),
                        skip = 1, col_names = c("gene_id","gene_symbol","gene_type"))

## Remove samples if there are fewer total fragment counts than there are possible features
## Somewhat arbitrary lower bound and shouldn't be an issue, but could catch a technical error
## if data isn't loading correctly (but you should see this in the mapping plot)
lowcounts <- colSums(assay(analysis$dds)) < nrow(geneInfoTab)
if (sum(lowcounts)>0){
  cat("Removing sample(s): ", names(lowcounts)[lowcounts], "due to low counts", sep="\n")
}
analysis$ddsDrop <- analysis$dds[,!lowcounts]
sample_order <- sample_order[!lowcounts]
sample_order <- droplevels(sample_order)
analysis$samplefileIDs <- analysis$samplefileIDs[!lowcounts]

rownames(analysis$ddsDrop) <- make.names(geneInfoTab$gene_symbol, unique=TRUE)
rownames(raw_counts) <- make.names(geneInfoTab$gene_symbol, unique=TRUE)
```

```{r vst_and_ddsDrop}
analysis$vst <- varianceStabilizingTransformation(analysis$ddsDrop,
                                                         blind = FALSE,
                                                         fitType = "parametric")
analysis$ddsDrop <- DESeq(analysis$ddsDrop, parallel = TRUE)
```

### Relative Log Expression (RLE) normalization

```{r rle, fig.width=12}
rawLogCounts <- log(counts(analysis$ddsDrop)[rowMins(counts(analysis$ddsDrop))>0,])
rawMedianLogs <- rowMedians(rawLogCounts)
rawLogRatios <- rawLogCounts - rawMedianLogs


normLogCounts <- log(DESeq2::counts(analysis$ddsDrop, normalized = TRUE)[rowMins(DESeq2::counts(analysis$ddsDrop))>0,])
normMedianLogs <- rowMedians(normLogCounts)
normLogRatios <- normLogCounts - normMedianLogs

analysis$RLE_raw <- RLE_plot(rawLogRatios, 'RLE Raw')
analysis$RLE_norm <- RLE_plot(normLogRatios, 'RLE Normalized')

print(analysis$RLE_raw)
print(analysis$RLE_norm)
```


### PCA {.active}

GEX PCA

```{r pca, fig.width=12, fig.height=8}
analysis$pca <- PCA_plot_from_config(analysis)
print(analysis$pca)
```

<!-- Overlaying age info on GEX PCA -->

<!-- ```{r, fig.width=12, fig.height=8} -->
<!-- pcaPlotGKT(analysis$vst, -->
<!--              intgroup = names(colData(analysis$vst)), -->
<!--              xpc = 1, ypc = 2) + -->
<!--     geom_point(aes(color = Age), -->
<!--                size = 5) + -->
<!--     labs(color='Age') + -->
<!--     (if (is.null(analysis$config$pcaMapping$label)) NULL else geom_text_repel(aes(label = .data[[analysis$config$pcaMapping$label]]), -->
<!--                     size = 4, hjust = 0.5, vjust = -0.5, alpha=0.5)) + -->
<!--     scale_x_continuous(expand = c(0.5,0)) + -->
<!--     theme_bw() + ggtitle(paste0(analysis$config$analysis," PCA")) + -->
<!--     theme(legend.key.width = unit(1.2, "cm")) + -->
<!--     theme(text = element_text(size=10)) -->
<!-- ``` -->

<!-- Overlaying collection date info on GEX PCA -->

<!-- ```{r, fig.width=12, fig.height=8} -->
<!-- pcaPlotGKT(analysis$vst, -->
<!--              intgroup = names(colData(analysis$vst)), -->
<!--              xpc = 1, ypc = 2) + -->
<!--     geom_point(aes(color = CollectionDate), -->
<!--                size = 5) + -->
<!--     labs(color='CollectionDate') + -->
<!--     (if (is.null(analysis$config$pcaMapping$label)) NULL else geom_text_repel(aes(label = .data[[analysis$config$pcaMapping$label]]), -->
<!--                     size = 4, hjust = 0.5, vjust = -0.5, alpha=0.5)) + -->
<!--     scale_x_continuous(expand = c(0.5,0)) + -->
<!--     theme_bw() + ggtitle(paste0(analysis$config$analysis," PCA")) + -->
<!--     theme(legend.key.width = unit(1.2, "cm")) + -->
<!--     theme(text = element_text(size=10)) -->
<!-- ``` -->


```{r write_tables, cache=TRUE, cache.lazy=FALSE, cache.rebuild=FALSE}
write_csv(analysis$sampleTable, path = paste0(analysis$config$analysis,".csv"), col_names = TRUE)
write_csv(as.data.frame(raw_counts)%>%rownames_to_column(var = "gene_id"),path = paste0("raw_count_",analysis$config$reference, "_", analysis$config$analysis,".csv"),col_names = TRUE)

ensembleRawCount <- do.call(cbind,sapply(analysis$sampleTable$FileID,
                                         function(sid){analysis$sampleSTARReads[[sid]][,"unstranded_count"][-c(1:4),]}))
rownames(ensembleRawCount) <- analysis$sampleSTARReads[[1]][-c(1:4),][["gene_id"]]

write_csv(as.data.frame(ensembleRawCount)%>%rownames_to_column(var = "EnsemblID"),
          path = paste0("raw_count_EnsemblID_", analysis$config$reference, analysis$config$analysis, ".csv"),col_names = TRUE)
analysis$rldDrop <- rlog(analysis$ddsDrop, blind = FALSE, fitType = "parametric")
analysis$assayRlogForGSEA <- assay(analysis$rldDrop)
# filter low/no-expression genes
analysis$assayRlogForGSEA <- analysis$assayRlogForGSEA[rowMeans(analysis$assayRlogForGSEA)>0,]
write_tsv(data.frame(Name = str_remove(rownames(analysis$assayRlogForGSEA), "[.].*"), Description = "na", analysis$assayRlogForGSEA), path = paste0("rlog_forGSEA_", analysis$config$analysis, ".txt"))
analysis$clsLinesGroup <- c(paste0(c(length(analysis$rldDrop$Group),length(unique(analysis$rldDrop$Group)),1), collapse = " "), paste0(c("#",unique(as.vector(analysis$rldDrop$Group))), collapse = " "), paste0(analysis$rldDrop$Group, collapse = " "))
write_lines(analysis$clsLinesGroup, path = paste0("Group_",analysis$config$analysis,".cls"))

```

## Pairwise histology group comparisons {.tabset}

```{r}
maxMinFilter <- function(object, intgroup = "condition", comp = c("ctrl","trmt"), thresh = 0)
{
  if (!all(intgroup %in% names(colData(object)))) {
    stop("the argument 'intgroup' should specify columns of colData(dds)")
  }
  intgroup.df <- as.data.frame(colData(object)[, intgroup, drop = FALSE])
  
  group <- if (length(intgroup) > 1) {
    interaction(lapply(intgroup, function(factorname) colData(ddsMoTime20034)[[factorname]]), drop = TRUE)
    #    factor(apply(intgroup.df, 1, paste, collapse = " : "))
  }
  else {
    colData(object)[[intgroup]]
  }
  if (!all(comp %in% levels(group))) {
    stop("the argument 'comp' should specify levels of intgroup")
  }
  #  if_else(rowMaxs(sapply(c("d0","d30"), function(lvl) rowMins(counts(ddsMoTime20034, normalize = TRUE)[,ddsMoTime20034$Timepoint == lvl]))) > 0,rowMeans2(counts(ddsMoTime20034, normalized = TRUE)),0)
  if_else(rowMaxs(sapply(comp, function(lvl) rowMins(counts(object, normalize = TRUE)[,group == lvl]))) > thresh,rowMeans2(counts(object, normalized = TRUE)),0)
}
```

```{r GSEA_setup}
## read in rlog file
#rlog_for_GSEA <- read.csv(paste0("rlog_forGSEA_", analysis$config$analysis, ".txt"), sep='\t')
## read GMT files
gmt.file <- list()
tcga_pathways <- readRDS('TCGA_pathway_genes.Rds')
cancer_pathways <- readRDS('pathways_cancer_genes.Rds')
gmt.file <- c(tcga_pathways, cancer_pathways)
  # for (a in file.path(analysis$config$rootDir, analysis$config$gseaFilesPath, analysis$config$gseaCollections)){#'mh.all.v0.3.symbols.gmt'
  #   gmt.file <- append(gmt.file, gmtPathways(file.path(a)))
  # }
```

```{r run_comparisons}

#### Add comparisons to get results for here as list of two lists: 
#### first list has the result names for coefficients to compare
#### second is a list that maps to the first list and says whether to run it as
#### results(analysis, name=<comparison>) or results(analysis, contrast=c(list(<comparisons)))
#### these can be very different and are important for multifactor designs
#### See example below. First list is the names/contrasts. If it is a contrast,
#### the two factors should be put as a list. A name can just be a character vector
####  my_comparisons <- list(                           # the whole thing is a list of lists
####            list('CD8Depleted.TimepointW1',         # the list of comparisons
####                  list('Timepoint_W7_vs_W0',        # the second member is a contrast, so a list of two names
####                       'CD8Depleted.TimepointW7'),
####                       'Timepoint_W1_vs_W0'),       # end of comparisons
####                  list('name', 'contrast', 'name')) # second list, saying if each member of the first 
####                                                    # list is a name or a contrast for the results call
#### see resultsNames(analysis$ddsDrop) for help choosing comparisons

# resultsNames(analysis$ddsDrop)

my_comparisons <- list(list(c('Groupplasmacytoid'),
                            c('Groupadenocarcinoma', 'Groupsarcomatoid', 'Groupsquamous', 'Groupsmall.cell')),
                       list(c('Groupadenocarcinoma'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupsquamous', 'Groupsmall.cell')),
                       list(c('Groupsarcomatoid'),
                            c('Groupplasmacytoid', 'Groupadenocarcinoma', 'Groupsquamous', 'Groupsmall.cell')),
                       list(c('Groupsquamous'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupadenocarcinoma', 'Groupsmall.cell')),
                       list(c('Groupsmall.cell'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupsquamous', 'Groupadenocarcinoma')))

pairwise_comparisons <- function(x, alpha=0.05){
  ## Generate DESeq2 results objects for contrasts of interest
  result <- results(analysis$ddsDrop, contrast = x, alpha = alpha,
                    listValues = c(1, -1/4), 
                    filter=maxMinFilter(analysis$ddsDrop, intgroup = 'Group',
                                        comp = c('adenocarcinoma', 'squamous',
                                                 'small cell', 'plasmacytoid',
                                                 'sarcomatoid')))
  # if (class(x) == 'list'){
  #   result <- results(analysis$ddsDrop, contrast=x, test='Wald', alpha=alpha)
  # } else if (class(x) == 'character'){
  #   result <- results(analysis$ddsDrop, contrast=x, test='Wald', alpha=alpha)
  outgroup <- paste(unlist(x[2]), collapse = '-')
  result@metadata$contrast <- paste(unlist(x[1]), outgroup, sep=' vs ')
  
  return(result)
}

## Get DESeq2 results for each comparison of interest
results <- mapply(pairwise_comparisons,my_comparisons)
wb <- createWorkbook()
## Convert these result objects to custom class with visualizations, etc
## option to write out each result as an individual sheet in the workbook
results <- lapply(results, my_results_transform, wb, write_out=FALSE)

## Name result objects from contrasts
names(results) <- lapply(results, function(x) x@metadata$contrast)


## Write all results to workbook 
write_output_workbook(wb, results, analysis)
saveWorkbook(wb, file=paste0(analysis$config$analysis, "_DESeq2_results.xlsx"), overwrite=T)
```

```{r prepare_results}
### you can add custom ggplot tracks to the outputs here, or edit datatables
### Input is the modifided DESeq2 object with the visualizations
output_results <- function(comp){
  # Just plaintext summary of the result object
  summary(results[[comp]])
  # Fixed window volcano plot
  volcanoPlot <- results[[comp]]@visualizations$volplot_fixed 
  # Interactive datatable of all DGE results for this comp
  DGEtable <- DT::datatable(data.frame(results[[comp]]) %>% select(baseMean, log2FoldChange, lfcSE, padj) %>% filter(padj < analysis$config$alpha) %>% mutate_if(is.numeric, round, 5) %>% na.omit() %>% rownames_to_column(), rownames = FALSE, caption='DESeq2 output for this comparison, significant genes only', filter='top', autoHideNavigation = TRUE)  %>% formatRound(c('baseMean','log2FoldChange', 'lfcSE'), digits=2)
  # Dotplot of top GSEA pathways
  gseaDotplot <- gsea_dotplot(head(results[[comp]]@gsea, 10), findtoppaths = FALSE, title='10 TCGA pathways')
  # Interactive table of GSEA results
  gseaTable <- DT::datatable(results[[comp]]@gsea[order(results[[comp]]@gsea$pval)] %>% select(pathway,pval,padj,ES,NES,size,), rownames = FALSE, filter='top', autoHideNavigation = TRUE, caption='GSEA enrichment results') %>% formatRound(c('pval','padj', 'ES','NES'), digits=4)
  
  return(list(volcanoPlot=volcanoPlot,DGEtable=DGEtable,gseaDotplot=gseaDotplot,gseaTable=gseaTable))
}
```

### Plasmacytoid vs other

```{r}
### call results outputs here with output_results(name_of_comparison)
### comparison names are based on the input lists earlier
comparison1 <- output_results('Groupplasmacytoid vs Groupadenocarcinoma-Groupsarcomatoid-Groupsquamous-Groupsmall.cell')
```

```{r, fig.height=6}
## kept separate to allow modified figure size
comparison1$volcanoPlot
```

```{r}
comparison1$DGEtable
```

```{r, fig.height=6}
comparison1$gseaDotplot
```

```{r}
comparison1$gseaTable
```

### Sarcomatoid vs other

```{r}
## Same as before with the next comparison. Repeat for however many comparisons you have
comparison2 <- output_results('Groupsarcomatoid vs Groupplasmacytoid-Groupadenocarcinoma-Groupsquamous-Groupsmall.cell')
```

```{r, fig.height=6}
comparison2$volcanoPlot
```

```{r}
comparison2$DGEtable
```

```{r, fig.height=6}
comparison2$gseaDotplot
```

```{r}
comparison2$gseaTable
```

### Small cell vs other

```{r}
## Same as before with the next comparison. Repeat for however many comparisons you have
comparison3 <- output_results('Groupsmall.cell vs Groupplasmacytoid-Groupsarcomatoid-Groupsquamous-Groupadenocarcinoma')
```

```{r, fig.height=6}
comparison3$volcanoPlot
```

```{r}
comparison3$DGEtable
```

```{r, fig.height=6}
comparison3$gseaDotplot
```

```{r}
comparison3$gseaTable
```

### Squamous vs other
```{r}
## Same as before with the next comparison. Repeat for however many comparisons you have
comparison4 <- output_results('Groupsquamous vs Groupplasmacytoid-Groupsarcomatoid-Groupadenocarcinoma-Groupsmall.cell')
```

```{r, fig.height=6}
comparison4$volcanoPlot
```

```{r}
comparison4$DGEtable
```

```{r, fig.height=6}
comparison4$gseaDotplot
```

```{r}
comparison4$gseaTable
```

### Adeno vs other
```{r}
## Same as before with the next comparison. Repeat for however many comparisons you have
comparison5 <- output_results('Groupadenocarcinoma vs Groupplasmacytoid-Groupsarcomatoid-Groupsquamous-Groupsmall.cell')
```

```{r, fig.height=6}
comparison5$volcanoPlot
```

```{r}
comparison5$DGEtable
```

```{r, fig.height=6}
comparison5$gseaDotplot
```

```{r}
comparison5$gseaTable
```

## Genes of interest {.tabset}

These heatmaps show the relative expression (regularized and log2 transformed) of genes. Individual samples are grouped by their histology.

### Top genes from pairwise comparisons {.active}

Taking the most significantly differentially up and down expressed genes in the aforementioned pairwise comparisons. Expression is normalized to the median of all samples for that gene. 

```{r, fig.width = 16, fig.height=10}
## creating an arbitrary genelist
filt_top_genes <- function(x) {
  a <- rownames(head(as.data.frame(x) %>% arrange(pvalue) %>% filter(log2FoldChange<0), 3))
  b <- rownames(head(as.data.frame(x) %>% arrange(pvalue) %>% filter(log2FoldChange>0), 3))
  return(c(a,b))
}
genelist <- unique(unlist(lapply(results, filt_top_genes)))

# top_genes <- data.frame(results(analysis$ddsDrop)) %>% filter(!is.na(pvalue)) %>% filter(abs(log2FoldChange)>log2(1.5))
# top_genes <- top_genes[order(top_genes$pvalue),]
# geneList <- (rownames(head(top_genes, 40))) 


heatmap_from_genelist2 <- function(geneList, data=analysis$rldDrop, title= ' '){
  ### makes a heatmap of the given list of genes, separating samples slightly by group variable
  ## data should be of type DESeqTransform
  hmap <- data[rownames(data) %in% geneList,
               !colnames(data) %in% analysis$config$dropSamples]
  hmap <- hmap[order(match(rownames(hmap), geneList)),]
  slices <- as.numeric(hmap@colData$Group)
  labels <- levels(hmap@colData$Group)[unique(slices)]
  baseline <- rowMedians(assay(hmap))
  hmap <- assay(hmap) - baseline
  Heatmap(hmap, show_row_names = TRUE, heatmap_legend_param = list(title="log2 fold\nexpression"),border="black",column_order=colnames(hmap),
          width = ncol(hmap)*unit(3, "mm"),
          height = nrow(hmap)*unit(3, "mm"),
          rect_gp = gpar(color="black"),
          column_title=title,
          cluster_rows=FALSE,
          column_names_gp = grid::gpar(fontsize = 8),
          row_names_gp = grid::gpar(fontsize = 8),
          column_split=slices,
          top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(col=NA),
                                                              #,fill = 2:length(unique(exp.design$Group))+1),
                                                              labels = labels, 
                                                              labels_gp = gpar(col = "black", fontsize = 10), labels_rot=90, height=unit(8,"cm"))), #
          column_gap=unit(2, "mm"),
          col=colorRamp2(c(-2, 0, 2), c("blue", "white", "red")))
}

## Pass in a genelist
par(mai=c(1,1,6,1))
heatmap_from_genelist2(genelist)
```

### Genes from TCGA publications 

A look at some marker genes identified in related bladder cancer literature [(1)](https://www.nature.com/articles/nature12965) [(2)](https://www.cell.com/cell/fulltext/S0092-8674(17)31056-5). 

```{r, fig.width = 16, fig.height=10}
genes <- c(
'FGFR3',
'miR-99a',
'miR-100',
'miR-125b',
'miR-145',
'ERBB2 ',#HER2
'KRT5',
'KRT6A',
'KRT14',
'EGFR',
'GATA3',
'FOXA1',
'UPK3A',
'ESR2',
'miR-220a',
'miR-220b',
'KRT20',
'PPARG',
'SNX31',
'UPK1A',
'UPK2',
'FGFR3',
'PGM5',
'DES',
'C7',
'p53',
'CD44',
'KRT',
'TP53',
'RB1',
'E2F3',
'CDH1',
'HER2',
'MDM2',
'CDKN1A',
# NONCODING RNAS lncRNAs
'DANCR',
'GAS5',
'MALAT1',
'NEAT1',
'NORAD',
'UCA1',
'CD274',
'PDCD1')
heatmap_from_genelist2(analysis$config$geneList)
```

```{r,  eval=FALSE}
png('tcga_all_genes_heatmap.png')
tmp <- heatmap_from_genelist2(unique(unlist(tcga_pathways)))
draw(tmp)
dev.off()
```


### Genes from Hippo Pathway

```{r, fig.width = 16, fig.height=10}
for (pathway in c('Hippo')){ #'WNT', 'RTK-RAS', 
  print(heatmap_from_genelist2(tcga_pathways[[pathway]], title = pathway))
}
```

### Manuscript genes

```{r}
genelist <- c('NECTIN4','CD274', 'FGFR3', 'UPK2', 'SNX31', 'PPARG', 'EGFR', 'RB1', 'ERBB2')
heatmap_from_genelist2(genelist)
```

### HPV genes

```{r}
genelist <- c('SLF1', 'ARHGAP4', 'RMI2', ' CRACDL', 'CENPK', 'CIDEB', 'DDB2', 'FAM111B', 'GRIN2C', 'LIG1', 'MGA', 'NEURL1B', 'NUSAP1', 'PHF19', 'POLD1', 'SP1', 'UBR7', 'WDR76')
heatmap_from_genelist2(genelist)
```

```{r}

genelist <- c('UBE2NL','APOBEC3H','RAD9B','GPR143', 'CMC4','SCML1', 'SH3GLB2','PGK1','THBD','NAMPT',  'CD46', 'NCSTN','MTMR2', 'P4HA2', 'NRP1', 'CXCL2','PTPN12','CAVIN3','EGLN1','TMEM87B', 'ERG', 'ESM1','STARD3NL', 'PEAR1', 'ARMT1', 'TYW1',  'ZNF134','TPGS1',  'SPRYD7', 'DLEU1',  'PLAC1',  'FAM124B',     'ROR1',     'ZNF418')
heatmap_from_genelist2(genelist)
```

```{r}
hpv_hppi <- read.csv('hpv_hppi.csv')
hpv_expression <- exp(analysis$rldDrop@assays@data@listData[[1]][rownames(analysis$rldDrop@assays@data@listData[[1]]) %in% hpv_hppi$Official.symbol,])
## Normalize by library size
hpv_expression <- sweep(hpv_expression,2,colSums(hpv_expression),`/`) * 1428

hpv_expression <- hpv_expression[hpv_hppi$Official.symbol,]

hpv_expression <- hpv_expression * hpv_hppi$Regression.coefficient

hpv_predictions <- stack(colSums(hpv_expression)) %>% 
  arrange(desc(values))
colnames(hpv_predictions) <- c('HPV_prediction', 'SampleID')
hpv_predictions <- hpv_predictions %>%
  left_join(analysis$sampleTable %>% select(SampleID, Group), by = 'SampleID')
```

```{r}
hpv_predictions %>% 
  filter(!grepl('Mixed', Group)) %>% 
  ggplot(aes(x = Group, group = Group, y = HPV_prediction, color = Group)) +
  geom_boxplot() + 
  theme_bw() +
  geom_jitter()
```



<!-- ### Requested Genes -->

<!-- ```{r, fig.height=6} -->
<!-- heatmap_from_genelist2(analysis$config$geneList) -->
<!-- ``` -->


## GSEA {.tabset}

### Pathways from publications

GSEA is conducted such that if the enrichment score is positive, the pathway is enriched in the first listed group, and if the enrichment score is negative, the pathway is enriched in the second group

```{r,  fig.height=8, fig.width=8}

# pathway_list <- c(
#   "WP_OXIDATIVE_STRESS_RESPONSE",
#   "GOBP_CELLULAR_RESPONSE_TO_DNA_DAMAGE_STIMULUS",
#   "GOBP_HISTONE_MODIFICATION",
#   "REACTOME_TRANSCRIPTIONAL_REGULATION_BY_TP53",
#   "GOBP_ALTERNATIVE_MRNA_SPLICING_VIA_SPLICEOSOME",
#   'BIOCARTA_HSWI_SNF_PATHWAY',
#   'REACTOME_COHESIN_LOADING_ONTO_CHROMATIN',
#   'GOBP_REGULATION_OF_CELL_CYCLE',
#   'GOBP_CHROMATIN_REMODELING'
# )
pathway_list <- names(gmt.file) %>% head(10)

gsea_plot_data <- lapply(names(results), function(x) {
  a <- results[[x]]@gsea
  a[a$pathway %in% pathway_list, ]
} )
names(gsea_plot_data) <- names(results)
gsea_plot_data <- data.table::rbindlist(gsea_plot_data, idcol='ID')
#tmp3 <- unlist(lapply(names(results), function(x) paste0(strsplit(results[[x]]@metadata$contrast, "_")[[1]][2:3], collapse=" over ")))
#tmp2$`.id` <- plyr::mapvalues(tmp2$`.id`, from= 1:10, to = names(results))
gsea_plot_data$pathway <- wrap_underscore_strings_balance(gsea_plot_data$pathway,36)
gsea_plot_data <- gsea_plot_data %>% separate(ID, c("ID"), extra = 'drop', sep=' ') %>% mutate(ID = substr(ID, 6, 100))

  ggplot(gsea_plot_data) + geom_point(aes(x=ID, y=pathway, size=-log(pval), color=NES)) +
    scale_color_gradient2(low="blue", mid="white",high="red", midpoint=0, breaks=c(-2,-1,0,1,2), limits=c(min(gsea_plot_data$NES,-1), max(gsea_plot_data$NES,1))) +
    theme_classic(11) +
    theme(panel.grid.major = element_line(colour = "grey92"),
          panel.grid.minor = element_line(colour = "grey92"),
          panel.grid.major.y = element_line(colour = "grey92"),#element_blank(),
          panel.grid.minor.y = element_line(colour = "grey92")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(x="Comparison", y="Gene set", color = "Normalized\nenrichment\nscore", size="Nom p-val", title="Pathways of interest") +  scale_radius(name="NOM p-val", range=c(1,10), breaks=-log10(c(0.1,0.01,0.001,0.0001)), labels=c(0.1,0.01,0.001,0.0001)) 

```

### Top pathways from this study

Taking the top 3 most significant pathways from each pairwise comparison in this study

```{r,  fig.height=10, fig.width=10}

gsea_plot_data <- lapply(names(results), function(x) {
  a <- results[[x]]@gsea
  head(a %>% arrange(pval), 3)$pathway
} )

gsea_plot_data <- lapply(names(results), function(x) {
  a <- results[[x]]@gsea
  a %>% filter(pathway %in% unlist(gsea_plot_data))
} )
names(gsea_plot_data) <- names(results)
gsea_plot_data <- data.table::rbindlist(gsea_plot_data, idcol='ID')
#tmp3 <- unlist(lapply(names(results), function(x) paste0(strsplit(results[[x]]@metadata$contrast, "_")[[1]][2:3], collapse=" over ")))
#tmp2$`.id` <- plyr::mapvalues(tmp2$`.id`, from= 1:10, to = names(results))
gsea_plot_data$pathway <- wrap_underscore_strings_balance(gsea_plot_data$pathway,36)
gsea_plot_data <- gsea_plot_data %>% separate(ID, c("ID"), extra = 'drop', sep=' ') %>% mutate(ID = substr(ID, 6, 100))



  ggplot(gsea_plot_data) + geom_point(aes(x=ID, y=pathway, size=-log(pval), color=NES)) +
    scale_color_gradient2(low="blue", mid="white",high="red", midpoint=0, breaks=c(-2,-1,0,1,2), limits=c(min(gsea_plot_data$NES,-1), max(gsea_plot_data$NES,1))) +
    theme_classic(11) +
    theme(panel.grid.major = element_line(colour = "grey92"),
          panel.grid.minor = element_line(colour = "grey92"),
          panel.grid.major.y = element_line(colour = "grey92"),#element_blank(),
          panel.grid.minor.y = element_line(colour = "grey92")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(x="Comparison", y="Gene set", color = "Normalized\nenrichment\nscore", size="Nom p-val", title="Top Pathways") +  scale_radius(name="NOM p-val", range=c(1,8), breaks=-log10(c(0.1,0.01,0.001,0.0001)), labels=c(0.1,0.01,0.001,0.0001))

```

## Requested plots

```{r requested_plot, include=TRUE}
requested_plot <- as.data.frame(log2(DESeq2::counts(analysis$ddsDrop, normalized = TRUE))) %>% filter(rownames(.) %in% c('NECTIN4' , 'CD274', 'TACSTD2', 'EFNB2'))
metadata <- t(analysis$sampleTable %>% select(c(SampleID, Group)) %>% filter(SampleID %in% colnames(requested_plot)))
colnames(metadata) <- metadata[1,]
requested_plot <- t(rbind(metadata, requested_plot))
requested_plot <- as.data.frame(requested_plot) %>% mutate_at(c('NECTIN4', 'TACSTD2', 'CD274', 'EFNB2'), as.numeric) %>% filter(!SampleID == 'Emory-102.2')
requested_plot <- melt(requested_plot, id.vars = c('SampleID', 'Group'))
ggplot(requested_plot, aes(x=Group, y=value, group=Group, color=Group)) + geom_boxplot()  + geom_jitter()+ facet_wrap(vars(variable), ncol = 4, strip.position = "top", scales = "free_x") +  theme_bw() +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +  scale_x_discrete(limits=c(
          'adenocarcinoma', 'Mixed (urothelial, adenocarcinoma)', 'sarcomatoid', 'Mixed (sarcomatoid, squamous)' ,'squamous', 'Mixed (Squamous, plamacytoid, micropap)', 'plasmacytoid', 'small cell')) + scale_color_manual(limits=c(
          'adenocarcinoma', 'Mixed (urothelial, adenocarcinoma)', 'sarcomatoid', 'Mixed (sarcomatoid, squamous)' ,'squamous', 'Mixed (Squamous, plamacytoid, micropap)', 'plasmacytoid', 'small cell'), values = c('red', 'pink', 'goldenrod', 'green', 'darkgreen', 'lightblue', 'blue', 'purple')) + labs(y='Normalized expression (log2 scaled)') #+ scale_y_log10() 
```

```{r}

```


```{r, eval=FALSE}

# for (result in results){
#   as.data.frame(result) %>% filter(rownames(.) %in% c('NECTIN4', 'TACSTD2', 'CD274', 'EFNB2')) %>% rownames_to_column('Gene')
# }
tmp <- lapply(results, function(result){as.data.frame(result) %>% filter(rownames(.) %in% c('NECTIN4', 'TACSTD2', 'CD274', 'EFNB2')) %>% rownames_to_column('Gene')})
tmp2 <- dplyr::bind_rows(tmp, .id='Comparison') 
tmp3 <-  strsplit(tmp2$Comparison, " ")
tmp2$Comparison <- lapply(tmp3, function(x){paste0(x[2],'_over_', x[3])}) 
write_csv(tmp2, 'requested_genes_stats.csv')
#tmp2 <- tmp2 %>% rownames_to_column('Gene')
```

## External data

```{r}
external_data <- dir('../p22162_Lara_Data/', pattern = '*gene_counts.tsv', recursive = TRUE, full.names = TRUE)

external_counts <- sapply(external_data, function(sid) {
  read_tsv(file.path(sid),
  col_names = c("gene_id","gene_name",	"gene_type",	"unstranded", "stranded_first",	"stranded_second",	"tpm_unstranded",	"fpkm_unstranded",	"fpkm_uq_unstranded"),
  skip=6)},
simplify = FALSE,
USE.NAMES = TRUE)

external_raw_counts <- do.call(cbind,sapply(external_data, function(sid){
  external_counts[[sid]][,'unstranded']
  }))
colnames(external_raw_counts) <- paste('external_sample', 1:362, sep = '_')
rownames(external_raw_counts) <- external_counts[[1]]$gene_name


external_samples <- data.frame(FileID=NA,
                               SampleID = colnames(external_raw_counts), 
                               SubjectNumber = colnames(external_raw_counts),
                               CollectionDate = NA,
                               Age = NA,
                               TumorRegions = NA,
                               MatchedNormal = 0,
                               CollectionSite = NA,
                               Group = 'Transitional cell carcinoma',
                               Batch = NA)
rownames(external_samples) <- external_samples$SampleID


full_counts <- merge(raw_counts, external_raw_counts, by = 'row.names')
rownames(full_counts) <- full_counts$Row.names
full_counts <- full_counts[,-1]
# external_raw_counts <- external_raw_counts[(rownames(external_raw_counts) %in% rownames(raw_counts)),]
# full_counts <- raw_counts[(rownames(raw_counts) %in% rownames(external_raw_counts)),]
# 
# external_raw_counts <- external_raw_counts[]
# 
full_sample_table <- rbind(analysis$sampleTable, external_samples)

analysis_external <- DESeqDataSetFromMatrix(full_counts,
  full_sample_table,
  design = as.formula(paste0("~ 0 + Group"))
)

analysis_external.vst <- varianceStabilizingTransformation(analysis_external,
                                                         blind = FALSE,
                                                         fitType = "parametric")
```

```{r, fig.width=12, fig.height=8}
  pcaPlotGKT(analysis_external.vst,
             intgroup = names(colData(analysis_external.vst)),
             xpc = 1, ypc = 2) +
    #  geom_path(aes(group = SubjectID)) 
    geom_point(aes(color = (if (is.null(analysis$config$pcaMapping$color)) NULL else .data[[analysis$config$pcaMapping$color]]),
                   shape = (if (is.null(analysis$config$pcaMapping$shape)) NULL else .data[[analysis$config$pcaMapping$shape]])),
               size = 3) +
    labs(color=analysis$config$pcaMapping$color, shape=analysis$config$pcaMapping$shape) +
    (if (is.null(analysis$config$pcaMapping$label)) NULL else geom_text_repel(aes(label = .data[[analysis$config$pcaMapping$label]]),
                    size = 4, hjust = 0.5, vjust = -0.5, alpha=0.5)) +
    scale_x_continuous(expand = c(0.5,0)) +
    theme_bw() + ggtitle(paste0(analysis$config$analysis," PCA")) +
    (if (!is.null(analysis$config$pcaMapping$path)){
      geom_path(aes(linetype=.data[[analysis$config$pcaMapping$path]]))
    }) +
    # geom_path(aes(linetype=(if (is.null(analysis$config$pcaMapping$path)) NULL else .data[[analysis$config$pcaMapping$path]]))) +
    theme(legend.key.width = unit(1.2, "cm")) +
    #labs(color="Weeks") +
    (if (!is.null(analysis$config$pcaMapping$ellipse)){
      stat_ellipse(aes(color=.data[[analysis$config$pcaMapping$ellipse]]), type="norm", level=0.67)
    })+
    theme(text = element_text(size=10)) -> PCA_external_plot #, arrow=arrow(ends="last", type="closed", length=unit(0.1, "inches")))
  
  PCA_external_plot
```

```{r}
analysis_external <- DESeq(analysis_external, parallel = TRUE)
```

```{r}
pairwise_comparisons_external <- function(x, alpha=0.05){
  ## Generate DESeq2 results objects for contrasts of interest
  result <- results(analysis_external, contrast = x, alpha = alpha, listValues = c(1, -1/5), filter=maxMinFilter(analysis$ddsDrop, intgroup = 'Group', comp = c('adenocarcinoma', 'squamous', 'small cell', 'plasmacytoid', 'sarcomatoid', 'Transitional cell carcinoma')))
  # if (class(x) == 'list'){
  #   result <- results(analysis$ddsDrop, contrast=x, test='Wald', alpha=alpha)
  # } else if (class(x) == 'character'){
  #   result <- results(analysis$ddsDrop, contrast=x, test='Wald', alpha=alpha)
  outgroup <- paste(unlist(x[2]), collapse = '-')
  result@metadata$contrast <- paste(unlist(x[1]), outgroup, sep=' vs ')
  
  return(result)
}
```

```{r}
my_comparisons_2 <- list(list(c('Groupplasmacytoid'),
                            c('Groupadenocarcinoma', 'Groupsarcomatoid', 'Groupsquamous', 'Groupsmall.cell', 'GroupTransitional.cell.carcinoma')),
                       list(c('Groupadenocarcinoma'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupsquamous', 'Groupsmall.cell', 'GroupTransitional.cell.carcinoma')),
                       list(c('Groupsarcomatoid'),
                            c('Groupplasmacytoid', 'Groupadenocarcinoma', 'Groupsquamous', 'Groupsmall.cell', 'GroupTransitional.cell.carcinoma')),
                       list(c('Groupsquamous'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupadenocarcinoma', 'Groupsmall.cell', 'GroupTransitional.cell.carcinoma')),
                       list(c('Groupsmall.cell'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupsquamous', 'Groupadenocarcinoma', 'GroupTransitiontal.cell.carcinoma')),
                       list(c('GroupTransitional.cell.carcinoma'),
                            c('Groupplasmacytoid', 'Groupsarcomatoid', 'Groupsquamous', 'Groupadenocarcinoma', 'Groupsmall.cell')))

results_2 <- mapply(pairwise_comparisons_external,my_comparisons_2)
##wb <- createWorkbook()
## Convert these result objects to custom class with visualizations, etc
## option to write out each result as an individual sheet in the workbook
results_2 <- lapply(results_2, my_results_transform, wb, write_out=FALSE)

## Name result objects from contrasts
names(results_2) <- lapply(results_2, function(x) x@metadata$contrast)
```

